An Innovative Storage Stack Addressing Extreme Scale Platforms and Big Data Applications

### 2014-04-22 paper.tex
major restructuring. title and abstract completely rewritten. Into largely changed. Rest of the paper reorganized to talk about each layer and then the cross-cutting ideas.

\begin{abstract}
The DOE Extreme-Scale Technology Acceleration Fast Forward Storage and IO Stack
project is going to have significant impact on storage systems design within
and beyond the HPC community. With phase 1 of the project complete, it is an
excellent opportunity to explore the complete design and how it will address
the needs of extreme scale platforms.  This paper not only provides a timely
summary of important aspects of the design specifications but also capture the
underlying reasoning that is not available elsewhere.

While many of the design elements are best practices from current research,
there are many new contributions as well as different integration approaches
offering different performance and quality of service guarantees. The specifics
of how the different layers interact offer a new approach for managing
scientific data for extreme scale platforms.

The integration of the HDF5 interface as the user interface gives a familiar
way to imteract with all of the new functionality while largely insulating 
end-users from the intracisies of dealing with the advanced functionality
offered by the stack.

The optional components aid in the scalability and simplifying implementation
across a wide range of platforms.

The data storage layer offers support both for local machine and data center
wide storage arrays as part of the design.

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform. To
address these limitations, a joint effort between the US Department of Energy's
Office of Advanced Simulation and Computing and Advanced Scientific Computing
Research commissioned an effort to develop a design and prototype for an IO
stack suitable for the extreme scale environment. This is a joint effort led by
Lawrence Livermore National Laboratory, with the DOE Data Management Nexus
leads Rob Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL. Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. Overriding many of the decisions during
this and any subsequent phases is the reality of budgets. Placing GBs of NVRAM
on every node, while a potentially advantageous approach, is not financially
feasible. With this in mind, the second phase will refine this design
incorporating fault recovery and other features missing from the first phase.

The overall design seeks to offer high availability, byte-granular,
multi-version concurrency control. Through the use of a copy-on-write style
mechanism, multiple versions of an object can be stored in potentially greatly
reduced space. It assumes the client interface will be through an IO library
affording a more complicated interface that offers richer functionality
requiring only minimal end-user code changes.  Managing most data access in a
platform-local layer rather than requiring writing to centralized storage will
better support the performance and energy requirements of extreme scale
integrated application workflows. At a more detailed view, the various layers
of the IO stack each contribute different functionality and performance
implications.

The basic architecture incorporates five layers. The top layer is a high level
IO library, such as the demonstration HDF-5 library~\cite{hdf5}. The intent is
to only have access to the storage stack through such an API to manage the
complexity of working with the lower layers. Below the user API is an IO
forwarding layer that redirects IO calls from the compute nodes to the IO
dispatching layer.  This IO forwarding layer is analogous to the function of
the IO nodes in a BlueGene machine. The next two layers have considerable
functionality. The IO dispatcher (IOD) serves as the primary storage interface
for the IO stack and is the only way to reach the persistent storage array in
lower layers. The Distributed Application Object Storage (DAOS) layer serves as
the persistent storage interface and is intended to be the foundation on which
everything else is built with no dependence on any technologies specified above
it. For example, the IOD layer and burst buffers are not required for DAOS to
operate properly. At the bottom is the Versioning Object Storage Device (VOSD).
It serves as the interface for storing objects of all types efficiently.

%%%%%%%%%%%%%%%%%%%%%%%%% <<<<<<<<<<<<<<<<

The ``I/O nodes'' handle 
requests forwarded by the scientific applications, integrate a tier of
solid-state devices to absorb the burst of random operations, and
organize/re-format the data so that transfers from the staging area to
the traditional parallel file system can be done more efficiently. It
also has the capacity of executing analysis on the fresh data that has
just been generated by simulation applications running at the compute
nodes. As the data arrives, re-organization and data preparation can be
applied in order to anticipate the execution of analytical tasks
{[}3{]}.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-arch.png}
\caption{Exascale Architecture. \label{exa-arch}}
\end{figure}

A common cluster configuration for this type of deployment is shown in
Figure \ref{exa-arch}. The designated I/O nodes (IONs) are connected to
the compute nodes (CNs) through the same fast fabric (eg. Infiniband),
while the connection to the external storage cluster is through a
secondary slower channel (eg. 10Gb ethernet). I/O staging thus
introduces a new layer in the application-to-storage stack, as shown in
Figure \ref{exa-stack} (left).

In current proposals, the stack is managed by middleware that seats
between the application and the parallel file system {[}2,4{]}. Since
most of the applications assume a POSIX interface to storage {[}5{]},
existing middleware either transparently handle the I/O operations,
appearing as regular POSIX calls to applications; or modify the I/O API
as little as possible in order to minimize the impact on production
codes.

The Fast Forward Storage I/O project {[}6{]} is aimed at merging the
features of existing middleware into a next-generation parallel file
system stack. This new storage API termed the I/O dispatcher (IOD) is
object-based, transactional, asynchronous and I/O-staging-aware (Figure
\ref{exa-stack} (right)).

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-stack.png}
\caption{Top-to-bottom view of the component stack. \label{exa-stack}}
\end{figure}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Arbitrarily Connected Graphs (ACG).
\item
  HDF5 extensions.
\item
  I/O Dispatcher (IOD).
\item
  Distributed Application Object Storage (DAOS).
\item
  Versioning Object Storage Device (VOSD).
\end{enumerate}

This will exemplify how applications make use of the Exascale stack. For
FF, support for
GraphLab 
and GraphBuilder will be prototyped, which are graph-processing
frameworks. GraphBuilder is a set of MapReduce tasks that extract,
normalize, partition and serialize a graph out of unstructured data, and
writes graph-specific formats into HDFS. These files are later consumed
by GraphLab, a vertex-centric, asynchronous execution engine that runs
directly on top of HDFS (i.e. non-MapReduce). The following illustrates
the architecture of both frameworks:

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.15]{images/graphlab-and-graphbuilder.png}
\caption{Figure 3. GraphLab and GraphBuilder stacks (taken from
{[}4{]}).}
\end{figure}

In order to make both work on top of the exascale stack, both have to be
modified. After these modifications are implemented, GraphBuilder will
be able to write the partitioned graph in (the newly proposed) HDF5
files which will thus be stored in the IOD nodes (or IONs) in a
parallel-optimized way. On the GraphLab side, HDF5-awareness will allow
the library to perform at high speeds by benefiting from the new
features (see next section). In general both frameworks will be modified
so that calls to HDFS-based formats are replaced by the proposed HDF5
ones. This is referred to as the HDF Adaptation Layer or HAL and will
provide, from the GraphBuilder/GraphLab point of view {[}5{]}:

\subsection{HDF5 extensions}\label{hdf5-extensions}

The extensions done to HDF5 allows an application to take full advantage
of the new exascale features. The additions comprise {[}6{]} (Figure 4):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Object-storage API based on HDF5 to support high-level data models.
  This exposes asynchronous, transactional semantics to the application,
  as well as end-to-end data integrity. It will also allows the usage of
  pointer data types, passing of hints down to the storage and support
  for asynchronous index building, maintenance and querying.
\item
  Virtual Object Layer (VOL) plugin that translates HDF5 API requests
  from applications to IOD API calls.
\item
  Function shipping from CN to IONs. This provides the application
  developer with the capacity of sending computation down to the IONs
  and get back results.
\item
  Analysis Shipping from CN to IONs or DAOS nodes. This is similar to 3
  but instead of returning the result over the network, it gets stored
  on the nodes and pointers to it are returned.
\end{enumerate}

In terms of the CN-ION communication model, a client/server architecture
is implemented {[}8{]}: every ION runs an IOFSL (I/O Function Shipping
Layer) server {[}9{]}; the IOFSL client is integrated into the HDF5
library which runs on each CN. A client can forward requests to any
number of IONs. Every I/O operation issued by HDF5 is asynchronously
shipped to the IOFSL server and asynchronously executed.

HDF5 {[}7{]} is a versatile data model containing complex data objects
and metadata. Its information set is a collection of datasets, groups,
datatypes and metadata objects. The data model defines mechanisms for
creating associations between various information items. The main
components of HDF5 are described below.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/hdf5.png}
\caption{Sample file illustrating the hierarchical structure of HDF5
files. \label{hdf5-sample}}
\end{figure}

\begin{itemize}
\item
  \textbf{File}: In the HDF5 data model the container of an HDF5 infoset
  is represented by a file. It is a collection of objects that also
  explains the relationship between them. Every file begins with a root
  group ''/'', which serves as the ''starting-point'' in the object
  hierarchy.
\item
  \textbf{Dataset}: HDF5 datasets are objects that represent actual data
  or content. Datasets are arrays which can have multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions), and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.
\item
  \textbf{Group}: A group is an explicit association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets or datatypes within it.
\item
  \textbf{Attribute}: Attributes are used for annotating datasets,
  groups, and datatype objects. They are datasets themselves, and are
  attached to existing objects they annotate.
\end{itemize}

For example, as shown in Figure \ref{hdf5-sample}, the file
Sample.h5 contains the root group which itself contains a
group G1 and two datasets, D1 and D2. Group
G1 contains a dataset D3. Attribute A1 is
linked to dataset D1. The objects and the relationships between
them can be represented as a B-tree, which is used internally by HDF5 to
index its objects.

@@

\subsection{Virtual Object Layer}\label{virtual-object-layer}

The Virtual Object Layer is an abstraction mechanism internal to the
HDF5 library {[}\textbf{???}{]}. As shown in Figure \ref{vol-arch} it is
implemented just below the public API. The VOL exports an interface that
allows writing plugins for HDF5, thereby enabling developers to store
objects in a format different from the default HDF5 file format. Plugin
writers provide an implementation for a set of functions that access
data on disk or on a parallel file system.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.20]{images/vol-arch.png}
\caption{Architectural view of the VOL abstraction mechanism.
\label{vol-arch}}
\end{figure}

\subsection{IOD VOL plugin}\label{iod-vol-plugin}

@@

\subsection{Fast Forward's I/O Dispatcher
API}\label{fast-forwards-io-dispatcher-api}

@@

{[}sharding. \label{sharding}{]}{[}sharding{]}

All these distinct alternatives result in having many different ways of
executing the same analysis task. In the future, we plan to extend our
design to incorporate an optimization phase that will prepare a task by
taking into account the distinct possible ways of executing and select
the most promising one. In our subsequent discussions, we consider only
data-movement optimization, i.e.~sending the analysis code as close as
possible to the data. In practice, this means we focus on identifying
sharding of datasets and execute code accordingly over the appropriate
shards.

%%%%%%%%%%%%%%%%%%%%%%%%% >>>>>>>>>>>>>>

The focus of this paper is primarily the IOD layer given the critical role it
has in the performance and functionality of the entire stack. Most of the key
features explored in this paper all have a strong presence in the IOD layer
motivating the focus of this examination.

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product. These insights were gained based on
personal conversations with the team
members~\cite{JohnBent,QuinceyKoziol,EricBarton} discussing some of the
potential challenges with the design as written. These ideas are presented to
give a fuller picture of where the project is going rather than dwelling on the
limitations of the published documents.

The rest of the paper is organized as follows. Section~\ref{sec:burst}
discusses some of the features of incorporating burst buffers as designed and
suggests some considerations and alternatives for the next generation of this
project. Section~\ref{sec:transactions} discusses the transactions approach
offered in the IOD layer and the corresponding epochs in the DAOS layer. It
also offers a comparison to the \DDT system given the very similar high-level
design and motivating use case.  Section~\ref{sec:summary} discusses the system
overall with recommendations on what design elements should be considered based
on broader issues with current HPC data centers. The paper is concluded in
Section~\ref{sec:conclusion} with a summary of the broad issues covered in the
paper.

\section{Burst Buffers}
\label{sec:burst}

@@

\subsection{Design Philosophy}
The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSD buffers to the IO nodes. The team
does acknowledge the limitations of this design and intend to ultimately focus
on spreading the IOD layer from the IO nodes into the compute area as well.
This is intended both to help address the limitations of the IO bandwidth and
compute capability of these few nodes for data processing, but also to take
advantage of new layers in the storage hierarchy. By incorporating NVRAM into
compute nodes, new options for buffering data prior to being moved to
centralized storage become available and addresses some of the concerns about
SSD performance. For example, including a small amount of Phase Change memory
into many or most compute nodes offers a way to move data outside of both the
compute and IO path for data and communication intensive operations. Other 
projects~\cite{zheng:2010:predata} have suggested this will have value, but the
cost will have to be considered as part of the overall platform budget. This
lessens the impact of some operators while offering additional options for
places to store data.

@@

\section{Transactions and Epochs}
\label{sec:transactions}

@@

\subsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container provides the single access context through which to access a
collection of objects. Transactions are the way that a series of modifications
to the objects within a container are treated atomically. Conceptually,
containers corresponds to a something akin to an HDF5 file in a traditional
file system. The objects in each container represent different data within a
file.  The three initially defined object types are key-value stores,
multi-dimensional arrays, and blobs.  The easiest way to understand these types
is to evaluate these from the perspective of an HDF-5 file, the initial user
interface layer. The key-value store represents a collection of attributes or
groups. The array represents a potentially multi-dimensional array.  The blob
represents a byte stream of arbitrary contents.  The fundamental difference
between an array and a blob is that the array has metadata specifying the
dimension(s). At the physical layer within the IO nodes, all of these objects
may be striped across multiple IO nodes.  Given this context, the transactions
come in two forms.

@@

\subsubsection{Design Philosophy}
Undocumented, but inherent in the design of these transactions is how faults
are detected. The initial design assumes the current Lustre fault detection
mechanism that can determine if a process or node is no longer reachable. This
detection happens at the DAOS layer and when a fault is detected, the rollback
process is pushed up to the IOD layer for all non-persisted or non-committed
transactions. This defines how a fault will be detected and what will trigger a
passive fault recovery (i.e., transaction abort).

@@

\subsection{DAOS Epochs}
The Epoch mechanism differs from transactions. Instead of focusing on when a
particular output is complete, an epoch represents incremental persisted copies
of a container. To simplify the mapping between an IOD transaction and the DAOS
epochs, when an IOD transaction is persisted to DAOS, the IOD transaction ID is
the used as the epoch ID. The key difference is that at the DAOS layer, some
transaction (epoch) IDs will not be represented since not all IOD transactions
are necessarily persisted.

@@

\subsubsection{Design Philosophy}
While the metadata design is not fully defined, there are a few things that
are intended. For example, there is a standard, well-known container that is
the system metadata. This includes the list of all other containers. This
container is treated like any other data in the system and striped as
appropriate. Unfortunately, this still couples the metadata to a single object
that must serialize access. If the metadata, including information about
striping and other data layout operations were separated completely from the
data path, more scalable throughput could be achieved. The real challenge of
this is introduced by the IOD, DAOS, and VOSD layers collectively. Each of
these requires some different metadata storage and the migration is transparent
to the user.  Supporting fully independent metadata with this model is
difficult. Serious thought on how to do this effectively outside the data path
should be considered for phase two.

>>>>

\begin{abstract}
The DOE Extreme-Scale Technology Acceleration Fast Forward Storage and IO Stack
project is going to have significant impact on storage systems design within
and beyond the HPC community. With phase 1 of the project complete, it is an
excellent opportunity to explore the complete design and how it will address
the needs of extreme scale platforms.  This paper not only provides a timely
summary of important aspects of the design specifications but also capture the
underlying reasoning that is not available elsewhere. With this information, we
encourate the broader community to understand the design, intent, and future
directions to foster discussion guiding phase 2 and the ultimate production
storage stack based on this work.

While many of the design elements are best practices from current research,
there are many new contributions as well as different integration approaches
offering different performance and quality of service guarantees. The specifics
of how the different layers interact offer a new approach for managing data for
extreme scale platforms and workloads.

For the current demonstration system, the HDF5 user API gives a familiar
way to interact with all of the new functionality while largely insulating 
end-users from the intracisies of dealing with the advanced functionality
offered by the stack.  The various layers below the user API include optional
components that may aid in scalability and simplify implementation across a
wide range of platforms.  The data storage layer offers support both
for local machine and data center wide storage arrays as part of the design.

@@

Current production HPC IO stack design is unlikely to offer sufficient features
and performance to adequately serve the needs of an extreme scale platform.
Adding to the complexity of the problem is the variety of Big Data problems
that are taking advantage of these platforms. Trying to address both of these
domains is a daunting problem.

To address these challenges, a joint effort between the US Department of
Energy's Office of Advanced Simulation and Computing and Advanced Scientific
Computing Research commissioned an effort to develop a design and prototype for
an IO stack suitable for the extreme scale environment called the Fast Forward
Storage and IO (FFSIO) project. This is a joint effort led by Lawrence
Livermore National Laboratory, with the DOE Data Management Nexus leads Rob
Ross and Gary Grider as coordinators and contract lead Mark Gary. The
participating labs are LLNL, SNL, LANL, ORNL, PNL, LBNL, and ANL.  Additional
industrial partners contracted include the Intel Lustre team, EMC, DDN, and the
HDF Group. This team has developed a specification
set~\cite{fastforward:2014:docs} for a future IO stack to address the
identified challenges. The first phase recently completed with a second phase
currently getting underway as of early 2014. The core focus of the first phase
was basic functionality and design. While an idealized potential system would
be the perfect target architecture, the reality of budgets has tempered many of
the decisions. For exmaple, extensive availability of NVRAN or SSDs on all of
the compute nodes is currently not economically feasible limiting some of the
potential design choices.  With this in mind, the second phase will refine this
design incorporating fault recovery and other features missing from the first
phase.

The complete design seeks to offer high availability, byte-granular,
multi-version concurrency control. Through the use of a copy-on-write style
mechanism reducing space and time requirements, multiple versions of an object
can be stored efficiently. By assuming the client interface will be through an
IO library, a more complicated interface that offers richer functionality can
be incorporated while requiring only minimal end-user code changes.  Managing
most data access in a platform-local layer rather than requiring writing to
centralized storage will better support the performance and energy requirements
of extreme scale integrated application workflows. At a more detailed view, the
various layers of the IO stack each contribute different functionality and
performance implications.

Overall, the architecture shifts from the idea of files and directories to
containers of objects. This shift avoids the bottlenecks related to the POSIX
files and directories structure such as the serialization of file creation, the
limitation and impact of the number of files in a directory, and the limited
semantics of a byte stream. Instead, the new interface focuses on high-level
data models and their properties and relationships. This concept permeates the
entire IO stack.

In addition to addressing the traditional scientific workload, this project
seeks to expand functionality to better support Big Data type applications. The
key idea is to support Arbitrary Connected Graphs (ACGs) such as those used in
Map-Reduce systems. Key system features are introduced to efficiently support
these computing models in addition to the typically bursty IO loads of more
traditional HPC applications.

The architecture incorporates five layers, some of which are potentially
optional. The top layer is a high level IO library, such as the demonstration
HDF-5 library~\cite{hdf5}. The intent is to only have access to the storage
stack through such an API to manage the complexity of working with the lower
layers and to enable advanced functionality that would require more direct user
intervention. This layer incorporate the necessary features for ACGs from a
end-user's perspective.

Below the user API is an IO forwarding layer that redirects IO calls from the
compute nodes to the IO dispatching layer.  This IO forwarding layer is
analogous to the function of the IO nodes in a BlueGene machine or the passive
data staging processes demonstrated
previously~\cite{nisar:2008:staging,Abbasi:2009:datatap}. The next two layers
have considerable functionality. The IO dispatcher (IOD) serves as the primary
storage interface for the IO stack and is the only way to reach the persistent
storage array in lower layers. Ideally, the IOD layer's functionality can be
optional based on available hardware. Much of the functionality offered at this
layer would shift either up or down the stack as discussed in detail below. The
Distributed Asynchronous Object Storage (DAOS) layer serves as the persistent
storage interface and is intended to be the traditional file system-like
foundation on which everything else is built with no dependence on any
technologies specified above it. For example, the IOD layer with or without
burst buffers is not required for DAOS to operate properly.  Instead, the DAOS
layer can handle all of the IO operations from the user API layer, albiet with
the potential performance penalty of manipulating the shared, persistent
storage array. At the bottom is the Versioning Object Storage Device (VOSD).
It serves as the interface for storing objects of all types efficiently. Think
of this layer as the physical disk interface layer. In terms of Lustre, this
would replace the file system on individual storage targets with an interface
friendlier to the containers of objects concept used in the higher layers.

Along with the analysis of the published design documents, a discussion of the
design philosophy representing the overall intent is presented. This
information represents information that may or may not have been written down,
but is the intent of ultimate product. These insights were gained based on
personal conversations with the team
members~\cite{JohnBent,QuinceyKoziol,EricBarton} understanding the current
design as written. These ideas are presented to
give a fuller picture of where the project is going rather than dwelling on the
limitations of the published documents.

The rest of the paper is organized as follows. A brief overview of related work
is presented first in Section~\ref{sec:related}. Section~\ref{sec:end-user}
discusses the programmatic interface end users will see when interacting with
the storage array. This will be discussed in the contect of the HDF-5 based
example library used for the functionality demonstration. Section~\ref{sec:iof}
briefly discusses the motivation and proposal for the IO forwarding layer.
Section~\ref{sec:iod} describes the IO Dispatcher layer and the broad
functionality it offers. This will detail the pieces of the layer that are
potentially optional and mention the cross-cutting features discussed in a
later, cross-cutting section. Section~\ref{sec:daos} discusses how the
persistent storage array interface itself functions. As with the IOD layer, the
disucssion of cross-cutting features will be mentioned, but disucssed more
fully in the cross-cutting section. The VOSD layer is disucssed in
Section~\ref{sec:vosd}. In particular, the mapping between the DAOS and VOSD
layers are explored as it pertains to the physical storage. Next is an
exploration of cross-cutting features like transactions and metadata management
in Section~\ref{sec:broader}. Since these and other features are spread across
multiple layers, it makes more sense to discuss them independently once an
understanding of the overall structure has been presented.  The paper is
concluded in Section~\ref{sec:conclusion} with a summary of the broad issues
covered in the paper.

\section{Related Work}
\label{sec:related}

Many projects over the last couple of decades have sought to address some
challenging aspect of parallel file system design. The recent rise of the ``Big
Data'' applications with different characteristic IO patterns have somewhat
complicated the picture. Extreme scale machines will be expected to handle both
the traditional simulation-related workloads as well as applications more
squarely in the Big Data arena. This will require some adjustments to the
underlying system for good performance for both scenarios.

The major previous work is really limited to full file systems rather than the
mountain of file system adjustments made over the years. The major players
related to this work include Ceph, PVFS, Lustre, GPFS, and Panasas.

EXPLAIN A BIT ABOUT HOW EACH CANNOT ADDRESS THE PROBLEMS WITH A CITATION FOR
EACH. DID I MISS A MAJOR FILE SYSTEM OF NOTE?

\section{End-User API Layer}
\label{sec:end-user}
Since the proposal specifies a high-level IO API will be the primary end-user
interface for programmatically interacting with the FFSIO stack, the team used
the HDF-5 API and leverated the Virtual Object Layer (VOL) for the initial
design and implementation demonstration. The VOL provides a way to intercept
HDF-5 API calls just below the API layer to implement alternative
functionality. For this project, rather than the default writing to disk in the
HDF-5 format, the VOL layer is used to interact with the IOD layer and the new
concepts it offers, such as transactions, completely hidden from the user.
This both offers a straightforward way for existing applications to test
against the FFSIO stack and to ensure proper semantics are maintained at the
lower levels. Some additional API functions have been introduced to support
new functionality, such as the ACG support for Hadoop-style applications.  The
additions comprise (Figure~\ref{fig:vol-arch}):

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt

\item
  Object-storage API based on HDF5 to support high-level data models.
  This exposes asynchronous, transactional semantics to the application,
  as well as end-to-end data integrity. It will also allows the usage of
  pointer data types, passing of hints down to the storage and support
  for asynchronous index building, maintenance and querying.

\item
  Virtual Object Layer (VOL) plugin that translates HDF5 API requests
  from applications to IOD API calls.

\item\label{item:fs}
  Function shipping from Compute Nodes (CN) to IO Nodes (ION). This provides
  the application developer with the capacity of sending computation down to the
  IONs and get back results.

\item
  Analysis Shipping from CN to IONs or DAOS nodes. This is similar
  to~\ref{item:fs} but instead of returning the result over the network, it
  gets stored on the nodes and pointers to it are returned.

\end{enumerate}

Function and Analysis Shipping are part of the cross-cutting features and are
discussed in Section~\ref{sec:fn-shipping}.

HDF5~\cite{hdf5} is a versatile data model containing complex data objects
and metadata. Its information set is a collection of datasets, groups,
datatypes and metadata objects. The data model defines mechanisms for
creating associations between various information items. The main
components of HDF5 are described below.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/hdf5.png}
\caption{Sample file illustrating the hierarchical structure of HDF5 files}
\label{fig:hdf5-sample}
\end{figure}

\begin{itemize}

\item
  \textbf{File}: In the HDF5 data model the container of an HDF5 infoset
  is represented by a file. It is a collection of objects that also
  explains the relationship between them. Every file begins with a root
  group ''/'', which serves as the ''starting-point'' in the object
  hierarchy.

\item
  \textbf{Dataset}: HDF5 datasets are objects that represent actual data
  or content. Datasets are arrays which can have multiple dimensions. A
  dataset is characterized by a dataspace and a datatype. The dataspace
  captures the rank (number of dimensions), and the current and maximum
  extent in each dimension. The datatype describes the type of its data
  elements.

\item
  \textbf{Group}: A group is an explicit association between HDF5
  objects. It is synonymous with directories in a file system. A group
  could contain multiple other groups, datasets or datatypes within it.

\item
  \textbf{Attribute}: Attributes are used for annotating datasets,
  groups, and datatype objects. They are datasets themselves, and are
  attached to existing objects they annotate.

\end{itemize}

For example, as shown in Figure \ref{fig:hdf5-sample}, the file Sample.h5
contains the root group which itself contains a group G1 and two datasets, D1
and D2. Group G1 contains a dataset D3. Attribute A1 is linked to dataset D1.
The objects and the relationships between them can be represented as a B-tree,
which is used internally by HDF5 to index its objects.

@@

\subsection{Virtual Object Layer}
\label{virtual-object-layer}

The Virtual Object Layer is an abstraction mechanism internal to the HDF5
library~\cite{hdf5}. As shown in Figure~\ref{fig:vol-arch} it is implemented
just below the public API. The VOL exports an interface that allows writing
plugins for HDF5, thereby enabling developers to handle the data in ways other
than necessarily writing to storage in an HDF5 format.  Plugin writers provide
an implementation for a set of functions that are trusted to provide the
proper semantics represented by the function for the new environment. For
example, data staging could be implemented in the VOL layer by replacing
writing to disk in the HDF5 format to sending data to a data staging area using
some messaging mechanism.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.20]{images/vol-arch.png}
\caption{Architectural view of the VOL abstraction mechanism}
\label{fig:vol-arch}
\end{figure}

\subsection{IOD VOL plugin}
\label{iod-vol-plugin}

@@

\section{IO Forwarding Layer}
\label{sec:iof}

This section will briefly describe the limited functionality of the IO
Forwarding layer and why it was included in the design.  It will not exceed
probably half of a column.

\section{IOD Layer}
\label{sec:iod}

\subsection{I/O Dispatcher API}
\label{sec:fast-forwards-io-dispatcher-api}

@@

All these distinct alternatives result in having many different ways of
executing the same analysis task. In the future, we plan to extend our
design to incorporate an optimization phase that will prepare a task by
taking into account the distinct possible ways of executing and select
the most promising one. In our subsequent discussions, we consider only
data-movement optimization, i.e., sending the analysis code as close as
possible to the data. In practice, this means we focus on identifying
sharding of datasets and execute code accordingly over the appropriate
shards.

\subsubsection{Sharding}
\label{sec:sharding}

The idea of sharding is to store a portion of a data set on a single storage
target. This is similiar in concept to the HDF-5 hyperslab. The FFSIO stack
supports sharding the data in the default or some other structured way as well
as ``re-sharding'' based on application needs. For example, reordering the
data so that a different dimension is the ``fast'' dimension may greatly
improve the performance of a subsequent data analytics task. A common scenario
where this is common is a Fortran code (column-major) writes data for a C code
(row-major) to analyze.

\subsection{I/O Nodes}
The ``I/O nodes'' handle requests forwarded by the scientific applications,
potentially integrate a tier of solid-state devices to absorb the burst of
random operations, and organize/re-format the data so that transfers to/from
the staging area from/to the traditional parallel file system can be done more
efficiently. It also has the capacity to executing analysis on the fresh data
that has just been generated by simulation applications running at the compute
nodes. As the data arrives, re-organization and data preparation can be applied
in order to anticipate the execution of analytical tasks.

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-arch.png}
\caption{Exascale Architecture}
\label{fig:exa-arch}
\end{figure}

A common cluster configuration for this type of deployment is shown in
Figure~\ref{fig:exa-arch}. The designated I/O nodes (IONs) are connected to the
compute nodes (CNs) through the same fast fabric (e.g., InfiniBand), while the
connection to the external storage cluster is through a secondary slower
channel (e.g., 10Gb Ethernet). I/O staging thus introduces a new layer in the
application-to-storage stack, as shown in Figure~\ref{fig:exa-stack} (left).

In current proposals, the stack is managed by middleware that sits between the
application and the parallel file system. Since most of the applications assume
a POSIX interface to storage, existing middleware either transparently handle
the I/O operations, appearing as regular POSIX calls to applications; or modify
the I/O API as little as possible in order to minimize the impact on production
codes.

The Fast Forward Storage I/O project is aimed at merging the features of
existing middleware into a next-generation parallel file system stack. This new
storage API termed the I/O dispatcher (IOD) is object-based, transactional,
asynchronous and I/O-staging-aware (Figure~\ref{fig:exa-stack} (right)).

\begin{figure}[htbp]
\centering
\includegraphics[scale=0.30]{images/exa-stack.png}
\caption{Top-to-bottom view of the component stack}
\label{fig:exa-stack}
\end{figure}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\itemsep1pt\parskip0pt\parsep0pt
\item
  Arbitrarily Connected Graphs (ACG).
\item
  HDF5 extensions.
\item
  I/O Dispatcher (IOD).
\item
  Distributed Asynchronous Object Storage (DAOS).
\item
  Versioning Object Storage Device (VOSD).
\end{enumerate}

\subsection{Burst Buffers}
\label{sec:burst}

@@

\noindent\textbf{Design Philosophy}

The burst buffers design, as presented in the IOD documents, limits the
placement of the function operators and SSD buffers to the IO nodes. The team
does acknowledge the limitations of this design and intend to ultimately focus
on spreading the IOD layer from the IO nodes into the compute area as well.
This is intended both to help address the limitations of the IO bandwidth and
compute capability of these few nodes for data processing, but also to take
advantage of new layers in the storage hierarchy. By incorporating NVRAM into
compute nodes, new options for buffering data prior to being moved to
centralized storage become available and addresses some of the concerns about
SSD performance. For example, including a small amount of Phase Change memory
into many or most compute nodes offers a way to move data outside of both the
compute and IO path for data and communication intensive operations. Other 
projects~\cite{zheng:2010:predata} have suggested this will have value, but the
cost will have to be considered as part of the overall platform budget. This
lessens the impact of some operators while offering additional options for
places to store data.

@@

\section{DAOS Layer}\label{sec:daos}
The Distributed Asynchronous Object Storage layer serves as the traditional
file-system interface layer for the storage devices.

\section{VOSD Layer}\label{sec:vosd}
The Versioning Object Storage Device layer operates as the layer for each
persistent storage device used to support the parallel storage array.

\section{Broader Design}
\label{sec:broader}

\subsection{Transactions and Epochs}
\label{sec:transactions}

@@

\subsubsection{IOD Transactions}
To understand how transactions are used in the IOD layer, some terminology and
concepts must be explained first. At the coarsest grain level is a container.
Each container provides the single access context through which to access a
collection of objects. Transactions are the way that a series of modifications
to the objects within a container are treated atomically. Conceptually,
containers corresponds to a something akin to an HDF5 file in a traditional
file system. The objects in each container represent different data within a
file.  The three initially defined object types are key-value stores,
multi-dimensional arrays, and blobs.  The easiest way to understand these types
is to evaluate these from the perspective of an HDF-5 file, the initial user
interface layer. The key-value store represents a collection of attributes or
groups. The array represents a potentially multi-dimensional array.  The blob
represents a byte stream of arbitrary contents.  The fundamental difference
between an array and a blob is that the array has metadata specifying the
dimension(s). At the physical layer within the IO nodes, all of these objects
may be striped across multiple IO nodes.  Given this context, the transactions
come in two forms.

@@

\noindent\textbf{Design Philosophy}

Undocumented, but inherent in the design of these transactions is how faults
are detected. The initial design assumes the current Lustre fault detection
mechanism that can determine if a process or node is no longer reachable. This
detection happens at the DAOS layer and when a fault is detected, the rollback
process is pushed up to the IOD layer for all non-persisted or non-committed
transactions. This defines how a fault will be detected and what will trigger a
passive fault recovery (i.e., transaction abort).

@@

\subsubsection{DAOS Epochs}
The Epoch mechanism differs from transactions. Instead of focusing on when a
particular output is complete, an epoch represents incremental persisted copies
of a container. To simplify the mapping between an IOD transaction and the DAOS
epochs, when an IOD transaction is persisted to DAOS, the IOD transaction ID is
the used as the epoch ID. The key difference is that at the DAOS layer, some
transaction (epoch) IDs will not be represented since not all IOD transactions
are necessarily persisted.

@@

\noindent\textbf{Design Philosophy}

While the metadata design is not fully defined, there are a few things that
are intended. For example, there is a standard, well-known container that is
the system metadata. This includes the list of all other containers. This
container is treated like any other data in the system and striped as
appropriate. Unfortunately, this still couples the metadata to a single object
that must serialize access. If the metadata, including information about
striping and other data layout operations were separated completely from the
data path, more scalable throughput could be achieved. The real challenge of
this is introduced by the IOD, DAOS, and VOSD layers collectively. Each of
these requires some different metadata storage and the migration is transparent
to the user.  Supporting fully independent metadata with this model is
difficult. Serious thought on how to do this effectively outside the data path
should be considered for phase two.
